---
title: "Data Science for Bioinformatics - Week 02"
author: "Palle Villesen"
date: "August 14, 2018"
output:
  html_document:
        theme: readable
---


### R for data science exercises #### 

Either you can do all the exercises in this document (to save your answers) or a new one.

The tutorial is from the free book "R for data science" written by the R Overlord Hadley Wickham

URL: http://r4ds.had.co.nz/index.html

 - Go through 4.    
 - Go through 5.   


#### Work on the mammals dataset (like last week)


```{r}

library(tidyverse)
mammals = read_csv(file = "../datasets/dataset.01.rsbl20150010supp1.csv")

```

# Gene expression in the mammals dataset

#### Q: Make a new variable called "expression" that is the log10(RPKM+1)


```{r}


```


#### Q: What is the overall distribution of the expression? (make a histogram)


```{r}


```

#### Q: What is the overall distribution of the expression but for each species separatedly?

```{r}


```

#### Q: What is the mean, median and stadard deviation of the expression for each Species?

```{r}


```

