---
title: "Data Science for Bioinformatics - week 10"
output: html_document
editor_options: 
  chunk_output_type: console
---



# Linear Models using the naked mole rat and the Mammals dataset 

The aim of this session is to go through the different steps of linear models fitting and model checking 

 * Fitting a series of linear models to data
 * Comparing models using F tests
 * understand differences in type I, II etc sum of squares
 * vizualizing the fit of a model to the data
 

# Load the data
In the book example, the goal is to compare energy expenditure of two castes of naked mole-rat while adjusting for differences in body mass using analysis of covariance (aka ANOVA).

```{r}
moleRat <- read.csv(url("http://www.zoology.ubc.ca/~schluter/WhitlockSchluter/wp-content/data/chapter18/chap18e4MoleRatLayabouts.csv"))
head(moleRat)
moleRatSorted <- moleRat[ order(moleRat$lnMass), ]
summary(moleRat)
```
## Model fitting

We fit severalmodels to the data, beginning with the model lacking an interaction term. We use the R function lm because both caste and mass are fixed effects. We first use a model with no interactions:

```{r}
moleRatNoInteractModel <- lm(lnEnergy ~ lnMass + caste, data = moleRatSorted) # fit model
moleRatSorted$predictedNoInteract <- predict(moleRatNoInteractModel) #save the predicted values in the data frame for vizualazing
```

Then, we fit a model with the interaction. Note that in the R function lm(), when we declare a general linear model of the form $Y = A*B$ it is a short hand for $Y = A + B + A*B$:

```{r}
moleRatFullModel <- lm(lnEnergy ~ lnMass * caste, data = moleRatSorted) 
moleRatSorted$predictedInteract <- predict(moleRatFullModel)
```

>Question: Visualize the fit of both models (no interaction or with interactions) to the data (see for instance the book figure)

```{r}
library(tidyverse)

ggplot(moleRatSorted) + geom_point(mapping = aes(lnMass, lnEnergy, shape = caste), size = 3) +
  geom_line(mapping = aes(lnMass, predictedNoInteract, linetype = caste), color = 'red', size = 1.2) + 
  geom_line(mapping = aes(lnMass, predictedInteract, linetype = caste), color = 'blue', size = 1.2)
```



# Testing for the effect using the appropriate F-ratio test 

This is were it gets slightly messy :-)- Reason is the design is not exaxtly balanced to it actually matterns how you construct F test (because the expected values of the F tests do change)
Here is a good read about the joys of the different types  (I, II etc) sum of squares:
http://www.uni-kiel.de/psychologie/rexrepos/posts/anovaSStypes.html


```{r}
#install.packages("car")
library(car) ## You need to install this library to get the type II or type III sum of squares
moleRatNoInteractModelTypeIII <- lm(lnEnergy ~ lnMass + caste, data = moleRat)
moleRatWIthInteractModelTypeIII <- lm(lnEnergy ~ lnMass * caste, data = moleRat)

#Question: Why are we testing type 3 here? The model did not account for interaction. Shouldn't it be type 2?
Anova(moleRatWIthInteractModelTypeIII, type = "III") # note "A" in Anova is capitalized

Anova(moleRatNoInteractModelTypeIII, type = "III") # note "A" in Anova is capitalized

```


## Model checking
You can use the  R diagnostic plots (just like in the case of ANOVA or regression) to visually check that the model assumptions are not strongly violated. 

```{r}
#install.packages("ggfortify")

plot(moleRatNoInteractModelTypeIII) ## Old style 
library(ggfortify) ## New style 
ggplot2::autoplot(moleRatNoInteractModelTypeIII)
ggplot2::autoplot(moleRatWIthInteractModelTypeIII)
```

# Reproducing the results obtained in the Mammals dn/ds and X versus autosome paper.
The goal here is to re-analyse the data using the model proposed by the authors of the paper.
The model fitted
```{r}
library(tidyverse)
mammals <- read_csv(file = "../datasets/dataset.01.rsbl20150010supp1.csv")

```

>Q Spend a bit of time asking yourslef what terms of the models are (continuous) covariates or diecrete ANOVA-like factors 

discrete: Species, chrMark
continuous: log(RPKM), Tau, GC3

>Q Refit the model used by the authors in their publication:

$$ log(dNdS) = Species+ chrMark + log(RPKM)+ Tau+ GC3 $$


To do that, adapt the naked mole rat R code and lm() and the Anova() function of the car package, to refit the model used by the authors and check your findings against the findings reported in Table 2 of the publication. Note that the authors use type II sum of squares to fit their model. 

```{r}
mam_mut = mammals %>% mutate(log_dNdS = log(dNdS), log_RPKM = log(RPKM))

mammalsNoInteractModel <- lm(log_dNdS ~ Species + chrMark + log_RPKM + Tau + GC3, data = mam_mut)
Anova(mammalsNoInteractModel, type = "II")

mammalsInteractModel <- lm(log_dNdS ~ Species + chrMark + log_RPKM + Tau + GC3
                           + Tau*log_RPKM + GC3 * log_RPKM + GC3*Tau 
                           + Species*log_RPKM + Species*Tau + Species*GC3, data = mam_mut)
Anova(mammalsInteractModel, type = "III")

```



# Exploring the effect of not accounting for confounding variables. 
A main conlusion of the authors is that most of the variation in dn/ds and the apparent "faster X" effect is due to the fact that - for other reasons- there is a tendency for genes located on the X to be lowly expressed (as measured by log(RPKM)).

>Q Fit a series of models that does not include the continuous covariates besides the species and chrMark factors.

Try four models, labelled S0, S1 S2 S3 (S for spurious ;-) ) :

$$ S_0: log(dNdS) =Species+ chrMark $$
```{r}
model1 <- lm(log_dNdS ~ Species + chrMark, data = mam_mut)
Anova(model1, type = "II")
```

$$ S_1: log(dNdS) =Species+ chrMark + Tau  $$
```{r}
model2 <- lm(log_dNdS ~ Species + chrMark + Tau, data = mam_mut)
Anova(model2, type = "II")
```


$$ S_2: log(dNdS) =Species+ chrMark+ GC_3 $$
```{r}
model3 <- lm(log_dNdS ~ Species + chrMark + GC3, data = mam_mut)
Anova(model3, type = "II")
```

$$ S_3: log(dNdS) =Species+ chrMark + log(RPKM)$$
```{r}
model4 <- lm(log_dNdS ~ Species + chrMark + log_RPKM, data = mam_mut)
Anova(model4, type = "II")
```

Examine the magnitude of this effect by finding out how big is the effect of the X versus Autosomes (the difference in means for that factors and what is the p-value attached to that effect) in models that do not include Tau log(RPKM) or GC3 as extra co-variates.


